A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-20 inst/manuscript/manuscript.Rmd
style guide updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for the approaches and then use a simulation study and an analysis of real mercury concentration data to numerically compare them. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, the model-based approaches often outperform design-based approaches, even for skewed data. This gap in performance is small when spatially balanced samples are used but large when non-spatially balanced samples are used.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
output: rticles::elsevier_article
---

@@

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:numstudy}, we use a simulation study to compare the performance of the approaches in a variety of scenarios. In Section \ref{application}, we compare the performance of the approaches on real data that contains mercury concentration in lakes from the contiguous United States. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

# Background {#sec:background}

@@

## Comparing Design-Based and Model-Based Approaches {#subsec:dvm_compare}

@@

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same data-generating stochastic process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Figure \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

@@

## Spatially Balanced Design and Analysis {#subsec:spb_design}

@@

## Finite Population Block Kriging

@@

# Simulation Study {#sec:numstudy}

@@

The average bias was nearly zero for all four combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for average bias in all 36 simulation scenarios are provided in the supplementary material.

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" as the baseline.
<!--- figure 2 -->
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}
The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Figure \ref{fig:figeff}, "Prop DE: 0" row), the four sampling-analysis combinations have approximately equal rMS(P)E. So using the GRTS sampling plan or a model-based analysis does not result in much, if any, loss in efficiency compared to IRS-Design when there is no spatial correlation. When there is spatial correlation (Figure \ref{fig:figeff}, "Prop DE: 0.5" and "Prop DE: 0.9" rows), GRTS-Model tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supplementary material.

@@

Figure \ref{fig:figconf} shows the 95% interval coverage for each of the four sampling-analysis combinations in the random location layout.
<!--- figure 3 -->
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}
Within each scenario, the sampling-analysis combinations tend to have fairly similar interval coverage. Coverage in the normal response scenarios was usually near 95%, while coverage in the lognormal response scenarios varied from from 90% to 95%. Coverage tended to always increase with the sample size. At a sample size of 200, all four sampling-analysis combinations had approximately 95% interval coverage in both response scenarios for all dependent error proportions.  These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for interval coverage in all 36 simulation scenarios are provided in the supplementary material.

# Application {#application}

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality for lakes in the contiguous United States. Specifically, we will analyze mercury concentration in lakes. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

Figure \ref{fig:merc} shows that mercury concentration is right-skewed, with most lakes having a low value of mercury concentration but a few having a much higher concentration. Mercury concentration exhibits some spatial correlation, with high mercury concentrations in lakes in the northeast and north central United States. The realized mean mercury concentration in the 986 lakes is 103.2 ng / g. TALK ABOUT EMPIRICAL SEMIVARIOGRAM IN THIS PARAGRAPH.

\begin{figure}
\centering
\begin{subfigure}{0.98\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption*{}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption*{}
  \label{fig:mercury_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/sv_plot.jpeg}
  \caption*{}
  \label{fig:sv_plot}
\end{subfigure}
\caption{Mercury concentration visualizations for the population (Hg) for 986 lakes in the contiguous United States. A spatial layout is in the top row, a histogram is in the bottom row and left column, and an empirical semivariogram is in the bottom row and right column.}
\label{fig:merc}
\end{figure}

@@

<!--- To better understand the dependence structure in mercury concentration, the empirical semivariogram can be visualized alongside a fit using the model-based approaches. The empirical semivariogram quantifies the halved squared differences (semivariance) among response values at different distances apart. If a process exhibits strong spatial dependence, the empirical semivariogram will have small values at small distances and large values at large distances. Figure \ref{fig:figsv} shows the empirical semivariogram quantifying the average semivariance for several distances.  Overlain onto Figure \ref{fig:figsv} is the estimated semivariance obtained using the covariance parameters from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides evidence that there is strong correlation in mercury concentration among the sites. -->

@@

This manuscript has a supplementary R package that contains all of the data and code used. Instructions for download at available at

# Supplementary Material {.unnumbered}

In the supplementary material, we provide tables presenting summary statistics for all 36 simulation scenarios.

@@

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1. 
  2.
  3.
  4.
  The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for the approaches and then use a simulation study and an analysis of real mercury concentration data to numerically compare them. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, the model-based approaches often outperform design-based approaches, even for skewed data. This gap in performance is small when spatially balanced samples are used but large when non-spatially balanced samples are used.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \doublespacing
output: rticles::elsevier_article
---

# Keywords {.unnumbered}

Design-based inference; Finite Population Block Kriging (FPBK); Generalized Random Tessellation Stratified (GRTS) algorithm; Model-based inference; Spatially balanced sampling; Spatial covariance;

@@

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we compare performance of the approaches wiht a simulation study and an analysis of real data that contains mercury concentration in lakes from the contiguous United States. In Section \ref{sec:results}, we present results from the simulation study and the analysis of mercury concentrations. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

## Background {#sec:background}

@@

### Comparing Design-Based and Model-Based Approaches {#subsec:dvm_compare}

@@

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same data-generating stochastic process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

@@

### Spatially Balanced Design and Analysis {#subsec:spb_design}

@@

### Finite Population Block Kriging

@@

# Materials and Methods {#sec:mm}

## Simulation Study {#sec:mm_sim}

@@

## Application {#sec:mm_app}

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use data from the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality for lakes in the contiguous United States [@USEPA2012NLA]. Specifically, we will analyze mercury concentration in lakes. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-analysis combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate

# Results {#sec:results}

## Simulation Study {#sec:r_sim}

The average bias was nearly zero for all four combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for average bias in all 36 simulation scenarios are provided in the supporting information.

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" as the baseline.
<!--- figure 2 -->
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}
The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Fig. \ref{fig:figeff}, "Prop DE: 0" row), the four sampling-analysis combinations have approximately equal rMS(P)E. So using the GRTS sampling plan or a model-based analysis does not result in much, if any, loss in efficiency compared to IRS-Design when there is no spatial correlation. When there is spatial correlation (Fig. \ref{fig:figeff}, "Prop DE: 0.5" and "Prop DE: 0.9" rows), GRTS-Model tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supporting information.

@@

Fig. \ref{fig:figconf} shows the 95% interval coverage for each of the four sampling-analysis combinations in the random location layout.
<!--- figure 3 -->
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}
Within each scenario, the sampling-analysis combinations tend to have fairly similar interval coverage. Coverage in the normal response scenarios was usually near 95%, while coverage in the lognormal response scenarios varied from from 90% to 95%. Coverage tended to always increase with the sample size. At a sample size of 200, all four sampling-analysis combinations had approximately 95% interval coverage in both response scenarios for all dependent error proportions.  These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for interval coverage in all 36 simulation scenarios are provided in the supporting information.

## Application {#sec:r_app}

Fig. \ref{fig:merc} shows that mercury concentration is right-skewed, with most lakes having a low value of mercury concentration but a few having a much higher concentration. Mercury concentration exhibits some spatial patterning, with high mercury concentrations in lakes in the northeast and north central United States. Fig. \ref{fig:merc} also shows the spatial dependence in mercury concentration via the empirical semivariogram. The empirical semivariogram can be used as a tool to visualize spatial dependence. It quantifes the halved squared differences (semivariance) among mercury concentration at different distances apart. When a process is spatially correlated, the semivariance tends to be smaller at small distances and larger at large distances. Together, the map, histogram, and semivariogram in Fig. \ref{fig:merc} suggest that mercury concentration is skewed and exhibits spatial dependence. Lastly we note that the realized mean mercury concentration in the 986 lakes is 103.2 ng / g.
\begin{figure}
\centering
\begin{subfigure}{0.98\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption*{}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption*{}
  \label{fig:mercury_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/sv_plot.jpeg}
  \caption*{}
  \label{fig:sv_plot}
\end{subfigure}
\caption{Mercury concentration visualizations for the population (Hg) for 986 lakes in the contiguous United States. A spatial layout is in the top row, a histogram is in the bottom row and left column, and an empirical semivariogram is in the bottom row and right column.}
\label{fig:merc}
\end{figure}

@@

@@

@@

# Conflict of Interest Statement {.unnumbered}

There are no conflicts of interest for any of the authors.

This manuscript has a supplementary R package that contains all of the data and code used in its creation. The supplementary R package is hosted on GitHub. Instructions for download at available at

# Supporting Information {.unnumbered}

In the supporting information, we provide tables presenting summary statistics for all 36 simulation scenarios.

# Author Contributions {.unnumbered}

All authors conceived the ideas; All authors designed methodology; MD and MH performed the simulations and analyzed the data; MD and MH led the writing of the manuscript; All authors contributed critically to the drafts and gave final approval for publication.

### 2021-12-20 inst/manuscript/manuscript.tex
style guide updates

\usepackage{bm} \usepackage{bbm} \usepackage{color} \DeclareMathOperator{\var}{{var}} \DeclareMathOperator{\cov}{{cov}} \usepackage{caption} \usepackage{subcaption}

@@

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference lie on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for the approaches
  and then use a simulation study and an analysis of real mercury
  concentration data to numerically compare them. We find that sampling
  plans that incorporate spatial locations (spatially balanced samples)
  perform better than sampling plans ignoring spatial locations
  (non-spatially balanced samples), regardless of whether design-based or
  model-based approaches were used to analyze the data. We also find that
  within sampling plans, the model-based approaches often outperform
  design-based approaches, even for skewed data. This gap in performance
  is small when spatially balanced samples are used but large when
  non-spatially balanced samples are used.
  \end{abstract}

@@

@@

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:numstudy}, we use a simulation study to
compare the performance of the approaches in a variety of scenarios. In
Section \ref{application}, we compare the performance of the approaches
on real data that contains mercury concentration in lakes from the
contiguous United States. And in Section \ref{sec:discussion}, we end
with a discussion and provide directions for future research.

\hypertarget{sec:background}{%
\section{Background}\label{sec:background}}

@@

\hypertarget{subsec:dvm_compare}{%
\subsection{Comparing Design-Based and Model-Based
Approaches}\label{subsec:dvm_compare}}

@@

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of
\emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. Prediction intervals are then
derived using assumptions of the data-generating stochastic process. If
we repeatedly generate response values from the same data-generating
stochastic process and select samples, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide thorough reviews of model-based
approaches for spatial data. In Figure \ref{fig:fig1}, we provide a
visual comparison of the design-based and model-based approaches (Ver
Hoef (2002) and Brus (2021) provide similar figures).

@@

\hypertarget{subsec:spb_design}{%
\subsection{Spatially Balanced Design and
Analysis}\label{subsec:spb_design}}

@@

\hypertarget{finite-population-block-kriging}{%
\subsection{Finite Population Block
Kriging}\label{finite-population-block-kriging}}

@@

\hypertarget{sec:numstudy}{%
\section{Simulation Study}\label{sec:numstudy}}

@@

The average bias was nearly zero for all four combinations in all 36
scenarios, so we omit a more detailed summary of those results here.
Tables for average bias in all 36 simulation scenarios are provided in
the supplementary material.

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout with ``IRS-Design'' as the baseline.

@@

The relative rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Figure
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Figure \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supplementary material.

@@

Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four sampling-analysis combinations in the random location layout.

@@

Within each scenario, the sampling-analysis combinations tend to have
fairly similar interval coverage. Coverage in the normal response
scenarios was usually near 95\%, while coverage in the lognormal
response scenarios varied from from 90\% to 95\%. Coverage tended to
always increase with the sample size. At a sample size of 200, all four
sampling-analysis combinations had approximately 95\% interval coverage
in both response scenarios for all dependent error proportions. These
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for interval
coverage in all 36 simulation scenarios are provided in the
supplementary material.

\hypertarget{application}{%
\section{Application}\label{application}}

The Environmental Protection Agency (EPA), states, and tribes
periodically conduct National Aquatic Research Surveys (NARS) in the
United States to assess the water quality of various bodies of water. We
will use the 2012 National Lakes Assessment (NLA), which measures
various aspects of lake health and water quality for lakes in the
contiguous United States. Specifically, we will analyze mercury
concentration in lakes. Although we know the true mean mercury
concentration values for the 986 lakes from the 2012 NLA, we will
explore whether or not we obtain an adequately precise estimate for the
realized mean mercury concentration if we sample only 100 of the 986
lakes.

Figure \ref{fig:merc} shows that mercury concentration is right-skewed,
with most lakes having a low value of mercury concentration but a few
having a much higher concentration. Mercury concentration exhibits some
spatial correlation, with high mercury concentrations in lakes in the
northeast and north central United States. The realized mean mercury
concentration in the 986 lakes is 103.2 ng / g. TALK ABOUT EMPIRICAL
SEMIVARIOGRAM IN THIS PARAGRAPH.

@@

This manuscript has a supplementary R package that contains all of the
data and code used. Instructions for download at available at

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

In the supplementary material, we provide tables presenting summary
statistics for all 36 simulation scenarios.

@@

@@

>>>>

\usepackage{bm} \usepackage{bbm} \usepackage{color} \DeclareMathOperator{\var}{{var}} \DeclareMathOperator{\cov}{{cov}} \usepackage{caption} \usepackage{subcaption} \usepackage{setspace} \doublespacing

@@

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
  \item
  \item
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    In this manuscript, we compare the approaches for finite population
    spatial data. We first provide relevant background for the approaches
    and then use a simulation study and an analysis of real mercury
    concentration data to numerically compare them. We find that sampling
    plans that incorporate spatial locations (spatially balanced samples)
    perform better than sampling plans ignoring spatial locations
    (non-spatially balanced samples), regardless of whether design-based
    or model-based approaches were used to analyze the data. We also find
    that within sampling plans, the model-based approaches often
    outperform design-based approaches, even for skewed data. This gap in
    performance is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used.
  \end{enumerate}
  \end{abstract}

@@

\hypertarget{keywords}{%
\section*{Keywords}\label{keywords}}
\addcontentsline{toc}{section}{Keywords}

Design-based inference; Finite Population Block Kriging (FPBK);
Generalized Random Tessellation Stratified (GRTS) algorithm; Model-based
inference; Spatially balanced sampling; Spatial covariance;

@@

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:mm}, we describe how we compare performance of
the approaches wiht a simulation study and an analysis of real data that
contains mercury concentration in lakes from the contiguous United
States. In Section \ref{sec:results}, we present results from the
simulation study and the analysis of mercury concentrations. And in
Section \ref{sec:discussion}, we end with a discussion and provide
directions for future research.

\hypertarget{sec:background}{%
\subsection{Background}\label{sec:background}}

@@

\hypertarget{subsec:dvm_compare}{%
\subsubsection{Comparing Design-Based and Model-Based
Approaches}\label{subsec:dvm_compare}}

@@

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of
\emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. Prediction intervals are then
derived using assumptions of the data-generating stochastic process. If
we repeatedly generate response values from the same data-generating
stochastic process and select samples, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide thorough reviews of model-based
approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual
comparison of the design-based and model-based approaches (Ver Hoef
(2002) and Brus (2021) provide similar figures).

@@

\hypertarget{subsec:spb_design}{%
\subsubsection{Spatially Balanced Design and
Analysis}\label{subsec:spb_design}}

@@

\hypertarget{finite-population-block-kriging}{%
\subsubsection{Finite Population Block
Kriging}\label{finite-population-block-kriging}}

@@

\hypertarget{sec:mm}{%
\section{Materials and Methods}\label{sec:mm}}

\hypertarget{sec:mm_sim}{%
\subsection{Simulation Study}\label{sec:mm_sim}}

@@

\hypertarget{sec:mm_app}{%
\subsection{Application}\label{sec:mm_app}}

The Environmental Protection Agency (EPA), states, and tribes
periodically conduct National Aquatic Research Surveys (NARS) in the
United States to assess the water quality of various bodies of water. We
will use data from the 2012 National Lakes Assessment (NLA), which
measures various aspects of lake health and water quality for lakes in
the contiguous United States (USEPA, 2012). Specifically, we will
analyze mercury concentration in lakes. Although we know the true mean
mercury concentration values for the 986 lakes from the 2012 NLA, we
will explore whether or not we obtain an adequately precise estimate for
the realized mean mercury concentration if we sample only 100 of the 986
lakes. For each of the four familiar sampling-analysis combinations
(IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate

\hypertarget{sec:results}{%
\section{Results}\label{sec:results}}

\hypertarget{sec:r_sim}{%
\subsection{Simulation Study}\label{sec:r_sim}}

The average bias was nearly zero for all four combinations in all 36
scenarios, so we omit a more detailed summary of those results here.
Tables for average bias in all 36 simulation scenarios are provided in
the supporting information.

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four approaches
from Table \ref{tab:designanalysis} using the random location layout
with ``IRS-Design'' as the baseline.

@@

The relative rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

Fig. \ref{fig:figconf} shows the 95\% interval coverage for each of the
four sampling-analysis combinations in the random location layout.

@@

Within each scenario, the sampling-analysis combinations tend to have
fairly similar interval coverage. Coverage in the normal response
scenarios was usually near 95\%, while coverage in the lognormal
response scenarios varied from from 90\% to 95\%. Coverage tended to
always increase with the sample size. At a sample size of 200, all four
sampling-analysis combinations had approximately 95\% interval coverage
in both response scenarios for all dependent error proportions. These
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for interval
coverage in all 36 simulation scenarios are provided in the supporting
information.

\hypertarget{sec:r_app}{%
\subsection{Application}\label{sec:r_app}}

Fig. \ref{fig:merc} shows that mercury concentration is right-skewed,
with most lakes having a low value of mercury concentration but a few
having a much higher concentration. Mercury concentration exhibits some
spatial patterning, with high mercury concentrations in lakes in the
northeast and north central United States. Fig. \ref{fig:merc} also
shows the spatial dependence in mercury concentration via the empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifes the halved squared
differences (semivariance) among mercury concentration at different
distances apart. When a process is spatially correlated, the
semivariance tends to be smaller at small distances and larger at large
distances. Together, the map, histogram, and semivariogram in Fig.
\ref{fig:merc} suggest that mercury concentration is skewed and exhibits
spatial dependence. Lastly we note that the realized mean mercury
concentration in the 986 lakes is 103.2 ng / g.

@@

@@

\hypertarget{conflict-of-interest-statement}{%
\section*{Conflict of Interest
Statement}\label{conflict-of-interest-statement}}
\addcontentsline{toc}{section}{Conflict of Interest Statement}

There are no conflicts of interest for any of the authors.

This manuscript has a supplementary R package that contains all of the
data and code used in its creation. The supplementary R package is
hosted on GitHub. Instructions for download at available at

\hypertarget{supporting-information}{%
\section*{Supporting Information}\label{supporting-information}}
\addcontentsline{toc}{section}{Supporting Information}

In the supporting information, we provide tables presenting summary
statistics for all 36 simulation scenarios.

\hypertarget{author-contributions}{%
\section*{Author Contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author Contributions}

All authors conceived the ideas; All authors designed methodology; MD
and MH performed the simulations and analyzed the data; MD and MH led
the writing of the manuscript; All authors contributed critically to the
drafts and gave final approval for publication.

@@

\leavevmode\hypertarget{ref-USEPA2012NLA}{}%
USEPA, 2012. National lakes assessment 2012.
https://www.epa.gov/national-aquatic-resource-surveys/national-results-and-regional-highlights-national-lakes-assessment.
